Multiple Output Regression

### 2014-02-11 paper/paper.tex
derivatives of inducing inputs

\noindent Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying similar integration we get,
\begin{align}
\log p(\vy)
\ge& \sum_{i=1}^P \sum_{n=1}^N
 \left( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta^{-1})
         - \frac{1}{2} \beta_i (w_i^2 \tilde{k}^g_{nn} + \tilde{k}_{inn})
         - \frac{1}{2} \trace \left( w_i^2 \S^g \mat{\Lambda}^g_n + \S_i \mat{\Lambda}_{in} \right)
\right) \\
&- \left( KL[q(\ug) || p(\ug)] + \sum_{i=1}^P KL[q(\vu_i) || p(\vu_i)] \right),
\end{align}
where 
\begin{align}
\tilde{\mu}_{in}
&= w_i k^g(\vx_n, \Zg)k^g(\Zg,\Zg)^{-1}\vm^g + k_i(\vx_n, \Zi)k_i(\Zi,\Zi)^{-1}\vm_i \\
\mat{\Lambda}^g_n
&= \beta k^g(\Zg,\Zg)^{-1} k^g(\Zg, \vx_n) k^g(\vx_n, \Zg) k^g(\Zg,\Zg)^{-1} \\
\mat{\Lambda}_{in}
&= \beta k_i(\Zi,\Zi)^{-1} k_i(\Zi, \vx_n) k_i(\vx_n, \Zg) k_i(\Zg,\Zg)^{-1}.
\end{align}

@@

>>>>

\noindent Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying similar integration we get,
\begin{align}
\nonumber
\log p(\vy)
\ge& \sum_{i=1}^P \sum_{n=1}^N
 \left( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta^{-1})
         - \frac{1}{2} \beta_i (w_i^2 \tilde{k}^g_{nn} + \tilde{k}_{inn})
         - \frac{1}{2} \trace \left( w_i^2 \S^g \mat{\Lambda}^g_n + \S_i \mat{\Lambda}_{in} \right)
\right) \\
&- \left( KL[q(\ug) || p(\ug)] + \sum_{i=1}^P KL[q(\vu_i) || p(\vu_i)] \right),
\end{align}
where 
\begin{align}
\tilde{\mu}_{in}
&= w_i k^g(\vx_n, \Zg)k^g(\Zg,\Zg)^{-1}\vm^g + k_i(\vx_n, \Zi)k_i(\Zi,\Zi)^{-1}\vm_i \\
\mat{\Lambda}^g_n
&= \beta k^g(\Zg,\Zg)^{-1} k^g(\Zg, \vx_n) k^g(\vx_n, \Zg) k^g(\Zg,\Zg)^{-1} \\
\mat{\Lambda}_{in}
&= \beta k_i(\Zi,\Zi)^{-1} k_i(\Zi, \vx_n) k_i(\vx_n, \Zg) k_i(\Zg,\Zg)^{-1}.
\end{align}

@@

\section{Appendix}
Here we derive the gradients of the lower bound wrt the  hyperparameters.
Consider the lower bound in Hensman et al (as a function of the hyperparameters):
\begin{align}
\nonumber
\calL
=& \log \Normal(\vy; \K_{NM} \K_{MM}^{-1} \vm, \beta^{-1}\I)
 - \frac{1}{2} \beta \trace \tilde{\K}
 - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\vm \vm^T + \S)) \right) \\ \nonumber
=& \underbrace{\log \Normal(\vy; \A \vm, \beta^{-1}\I)}_{\calL_1}
 - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2}
 - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3} \\  
&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\vm \vm^T + \S)) \right)}_{\calL_4},
\end{align}
where $\A = \K_{NM} \K_{MM}^{-1}$.
Notice that we have re-written the sum of individual terms in matrix form which will make the derivation and also computation easier.

\subsection{Derivative of the Noise Hyperparameter}
The derivative of the noise hyperparameter $\beta$ is easily computed as:
\begin{align}
\frac{d \calL}{d\beta} = \frac{N}{2\beta} - \frac{1}{2} (\vy - \A\vm)^T (\vy - \A\vm) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
\end{align}
\subsection{Derivatives of the Covariance Hyperparameters}  
To simplify the math, we utilize the matrix $\A$ defined above.
Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
\begin{align}
\frac{d \A}{dt} = \left(\frac{d\K_{NM}}{dt} - \A \frac{d \K_{MM}}{dt}\right)\K_{MM}^{-1}.
\end{align}
The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
\begin{align}
\frac{d \calL_1}{dt} &= \beta (\vy - \A\vm)^T \frac{d \A}{dt} \vm \\
\frac{d \calL_2}{dt} &= \frac{1}{2}\beta \trace \left(\frac{d\K_{NN}}{dt} - \A \frac{d\K_{MN}}{dt} - \frac{d\A}{dt} \K_{MN}\right) \\
\frac{d \calL_3}{dt} &= \beta \trace \left(\A \S \frac{d\A^T}{dt} \right) \\
\frac{d \calL_4}{dt} &= \frac{1}{2}  \trace \left(\K_{MM}^{-1} \frac{d \K_{MM}}{dt}\right) - \frac{1}{2} \trace \left(\K_{MM}^{-1} \frac{d \K_{MM}}{dt} \K_{MM}^{-1} (\vm \vm^T + \S) \right) 
\end{align}
The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.

\subsection{Derivatives of the Inducing Inputs}
To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\vz_m\}_{m=1}^M$ are also parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
Hence the derivative wrt a dimension of an inducing input, i.e. $z_{mj}$, is the same as that of $\frac{d \calL}{dt}$.

%Since $MD$ parameters are needed for the inducing inputs, it appears that the derivatives of all inducing inputs would require $\calO(MD \times M^3)$ in computation.
%However, this complexity can actually be reduced to $\calO(DM^3)$ using the following lemma. \\
%
%\noindent \textbf{Lemma} Let $A, B$ be two matrices of size $N \times M$ and $M \times N$, respectively. Furthermore, $B$ has the property that only one of its rows or columns is non-zero. The complexity of $\trace(AB)$ is only $\calO(N)$. \\
%
%\noindent \textbf{Proof} Let $m <= M$ be the non-zero row of $B$. We have
%\begin{align}
%\trace (AB) = \sum_{i=1}^N \sum_{j=1}^M A_{ij} B_{ji} = \sum_{i=1}^N A_{im} B_{mi},
%\end{align}
%which clearly takes $\calO(N)$. It is easy to see that the lemma also holds when $B$ is symmetric and only one of its row and the corresponding column is non-zero. \\
%
%\noindent To exploit the property in the Lemma, we re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):

\noindent We re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):
\begin{align}
\nonumber
% dL1
\frac{d \calL_1}{dt}
 &= \beta \trace (\vy - \A\vm)^T \left(\frac{d\K_{NM}}{dt} -  \A \frac{d \K_{MM}}{dt}\right)\K_{MM}^{-1} \vm \\
&= \beta \trace \K_{MM}^{-1} \vm (\vy - \A\vm)^T \frac{d\K_{NM}}{dt} 
-\beta \trace \K_{MM}^{-1} \vm (\vy - \A\vm)^T \A \frac{d \K_{MM}}{dt} \\
%dL2
\frac{d \calL_2}{dt}
&= - \beta \trace \A^T \frac{d \K_{NM}}{dt}
 + \frac{1}{2} \beta \trace \A^T \A \frac{d \K_{MM}}{dt}  \\
% dL3
\frac{d \calL_3}{dt}
&= \beta \trace \K_{MM}^{-1} \S \A^T \frac{d \K_{NM}}{dt}
 - \beta \trace \K_{MM}^{-1} \S \A^T \A \frac{d\K_{MM}}{dt}\\
% dL4
\frac{d \calL_4}{dt}
 &= \frac{1}{2}  \trace \K_{MM}^{-1} \frac{d \K_{MM}}{dt}
  - \frac{1}{2} \trace \K_{MM}^{-1} (\vm \vm^T + \S) \K_{MM}^{-1} \frac{d \K_{MM}}{dt} 
\end{align}
From the above equations we see that,
\begin{align}
\frac{d\calL}{dt} = \trace \D_1 \frac{d\K_{NM}}{dt} + \trace \D_2 \frac{d \K_{MM}}{dt},
\end{align}
where 
\begin{align}
\D_1 =& \beta \K_{MM}^{-1} \vm (\vy - \A\vm)^T
 + \beta \A^T
 - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
\D_2 =& -\beta \trace \K_{MM}^{-1} \vm (\vy - \A\vm)^T \A
 - \frac{1}{2} \beta \A^T \A
  + \beta \K_{MM}^{-1} \S \A^T \A	 \\ 
  &-\frac{1}{2} \K_{MM}^{-1} + \frac{1}{2} \K_{MM}^{-1} (\vm \vm^T + \S) \K_{MM}^{-1}
\end{align}
Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of $\trace \D_2 \frac{d \K_{MM}}{dt}$ is $\calO(M)$ due to the fact that only one row (and its corresponding column) of $\frac{d \K_{MM}}{dt}$ is non-zero.
This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.
